### 2018-07-01 paper/5-implementations.tex
Work on S5.2: simplify the model of Excel

Our \Excel's model is a composition of the \hs{restarting} scheduler and the
\hs{approximateRebuilder}, as shown in Fig.~\ref{fig-excel-implementation}.
The persistent build information \hs{ExcelInfo}~\hs{k} is a triple:

\begin{itemize}
    \item A set of dirty keys \hs{Set}~\hs{k}, i.e. all cells whose contents
          has changed since the previous build, including input cells and
          formulas.
    \item A map \hs{Map}~\hs{k}~\hs{[@@k]} of known upper bounds of
          dependencies. If there is an entry for a key, it is a conservative
          approximation of its dependencies (e.g. the dependencies of the
          formula \cmd{IF(C1=1,B2,A2)} are approximated by the list
          \hs{[}\cmd{C1}\hs{,}\cmd{B2}\hs{,}\cmd{A2}\hs{]}). Otherwise, there is
          no upper bound (e.g. because the key uses an \cmd{INDIRECT} reference)
          and the key is assumed to be dirty.
    \item A calculation \hs{Chain} recorded in the previous build
          (\S\ref{sec-background-excel}), represented by a list of keys \hs{[@@k]}.
\end{itemize}

The \hs{approximateRebuilder} uses the first two elements of the above triple
to decide whether a given \hs{key} is \hs{dirty} and hence needs to be rebuilt.
The logic is similar to that of \Make's \hs{modTimeRebuilder}, but instead of
relying on file modification times, \Excel stores the information about dirty
keys more directly. Furthermore, \Excel's dependency approximation makes it
possible to build tasks with dynamic dependencies: a \hs{key} is declared
\hs{dirty} if any of its approximate dependencies is in the \hs{dirtyKeys} set,
or if there is no known upper bound on its dependencies, i.e. if the lookup in
the map \hs{deps}~\hs{::}~\hs{Map}~\hs{k}~\hs{[@@k]} fails. Note that when a
\hs{key} is rebuilt it is added to the set \hs{dirtyKeys} to trigger further
rebuilds.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let value = getValue key store
                newTask :: Task (MonadState ir) k (Either k v)
                newTask = try $ rebuilder key value task
                fetch :: k -> State ir (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            result <- liftStore (run newTask fetch)
            case result of
                Left dep -> go done $ dep : filter (/= dep) ks ++ [key]
                Right newValue -> do
                    modify $ updateValue key value newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (implementation omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-5mm}
\end{figure}

\Excel's \hs{restarting} scheduler processes keys in the order specified by the
calculation \hs{chain}. During the build, it constructs a \hs{newChain} for the
next build and maintains a set of keys \hs{done} that have been processed. For
each non-input \hs{key}, the scheduler tries to rebuild it using a partial
\hs{fetch} callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}.
The callback is defined to fail with \hs{Left}~\hs{dep} when asked for the value
of a dependency \hs{dep} that has not yet been processed (and hence may
potentially be dirty); otherwise it returns the current value of the dependency
by looking it up in the \hs{store}.

After the \hs{newTask} is executed with \hs{runState} there are two cases to
consider:

\begin{itemize}
    \item The \hs{newTask} has failed, because one of its dependencies \hs{dep}
    has not yet been processed. This indicates that the calculation \hs{chain}
    from the previous build is incorrect and needs to be adjusted by making sure
    it contains \hs{dep} and moving the \hs{key} to the end, so that its build
    can be restarted after the \hs{dep} is ready.
    \item The \hs{newTask} succeeded. The resulting \hs{newValue} and
    \hs{newInfo} are written to the store, the \hs{key} is marked as \hs{done},
    and \Excel continues to build the \hs{chain}.
\end{itemize}

Note that the task returned by the \hs{rebuilder} expects a total callback
function and cannot be directly executed with the partial callback \hs{fetch}.
We fix the mismatch with the function \hs{try} that relies on the standard
monad transformer \hs{ExceptT} from the \cmd{transformers} library.

>>>>

Our model of \Excel uses the \hs{restarting} scheduler and the
\hs{dirtyBitRebuilder}, see Fig.~\ref{fig-excel-implementation}. The persistent
build information \hs{ExcelInfo}~\hs{k} is a pair: (i) a map
\hs{k}~\hs{->}~\hs{Bool} associating a dirty bit with every key, and (ii) a
calc \hs{Chain}~\hs{::}~\hs{[@@k]} recorded in the previous build
(\S\ref{sec-background-excel}).

The \hs{dirtyBitRebuilder} is very simple: if the \hs{key}'s dirty bit is set,
we \hs{run} the \hs{task} to rebuild the \hs{key}; otherwise we return the
current \hs{value} as is. The external system, which invokes \Excel's build
engine, is required to provide a transitively closed set of dirty bits.
Therefore, unlike \Make's \hs{modTimeRebuilder}, the \hs{dirtyBitRebuilder} does
not need to modify \hs{i} to trigger rebuilds of dependent keys.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let newTask :: Task (MonadState ir) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State ir (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            result <- liftStore (run newTask fetch) -- liftStore is defined in Fig. 7
            case result of
                Left dep -> go done $ dep : filter (/= dep) ks ++ [key]
                Right newValue -> do
                    modify $ putValue key newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (implementation omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-5mm}
\end{figure}

\Excel's \hs{restarting} scheduler processes keys in the order specified by the
calc \hs{chain}. During the build, it constructs a \hs{newChain} for the next
build and maintains a set of keys \hs{done} that have been processed. For each
non-input \hs{key}, the scheduler tries to rebuild it using a partial \hs{fetch}
callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}. The callback
is defined to fail with \hs{Left}~\hs{dep} when asked for the value of a
dependency \hs{dep} that has not yet been processed (and hence may potentially
be dirty); otherwise it returns the current value of the dependency by looking
it up in the \hs{store}.

After the \hs{newTask} is executed (with the help of \hs{liftStore}) there are
two cases to consider:

\begin{itemize}
    \item The \hs{newTask} has failed, because one of its dependencies \hs{dep}
    has not yet been processed. This indicates that the calculation \hs{chain}
    from the previous build is incorrect and needs to be adjusted by moving the
    \hs{dep} in front of the \hs{key}, so that we can restart building the
    \hs{key} after the \hs{dep} is ready.
    \item The \hs{newTask} succeeded. The resulting \hs{newValue} is written to
    the store, the \hs{key} is marked as \hs{done}, and \Excel continues to
    build the rest of the \hs{chain}.
\end{itemize}

Note that the task returned by the \hs{rebuilder} expects a total callback
function and cannot be directly executed with the partial callback \hs{fetch}.
We fix the mismatch with the function \hs{try} that relies on the standard
monad transformer \hs{ExceptT} from the \cmd{transformers} library. We also
need the \hs{liftChain} helper function, whose implementation we omit since it
is analogous to \hs{liftStore} defined in Fig.~\ref{fig-make-implementation}.
